### 2023-04-26 chapter2/chapter2.tex
Updated introduction and discussion outlines
Converted introduction paragraph 2 outline to paragraph

% II. Proteome-wide patterns of conserved features linked to function
%    B. Despite the prevalence and importance of these regions in eukaryotic proteomes, associating them with specific molecular functions or biological processes remains a significant challenge.
%    C. The sequences of IDRs are generally poorly conserved, so traditional bioinformatics approaches which rely on the strict conservation of amino acid sequences to identify homologous proteins and transfer annotations between biological systems are largely unsucessful when applied to IDRs.

@@

%    A. However, several recent studies have demonstrated evidence that IDRs are constrained to preserve ``distributed features'' such as flexibility, chemical composition, net charge, or charge distribution.
%       1. Because many sequences can yield a region with a specific composition or charge, this mode of constraint uncouples an IDR's fitness from its strict sequence of amino acid residues. the sequences of IDRs can rapidly diverge while still preserving these structural or functional properties.
%       2. Furthermore, in contrast to folded regions whose precise contacts and packing geometries are easily disrupted by point mutations, distributed features are robust to small changes at the sequence level. Thus, deleterious mutations are easily corrected by compensatory mutation elsewhere in the region. Together, these loose restrictions on the sequences of IDRs allow them to rapidly diverge despite preserving their functions.
%    B. Additionally other studies have shown importance of short linear motifs in IDRs
%        1. Can be identified as short segments of conservation in otherwise highly diverged alignments of IDRs
%        2. However, these segments comprise a small fraction of disordered regions, so
% III. Previously, many of these studies were limited to ... However, a recent study first to demonstrate patterns of selection in IDRs across an entire proteome using a set of 82 ``molecular features''

@@

%    E. Subsequent studies have expanded on these findings by using techniques from machine learning and statistics the information from evolutionary signatures to predict the function of IDRs. and learn disorder-associated features in a self-supervised manner.

@@

% III. Disorder score analyses
%    A. Comparison of features and disorder scores acts as a kind of ``mutational scanning'' experiment [citation to Eddy?]
%        1. Contrasts of features and disorder scores shows the disorder predictor is sensitive to most features used in this analyses
%        2. However, highest correlations to expected features like fraction_disorder and hydropathy
%        3. Interestingly, negatively correlated with wf_complexity. Many disordered regions have low levels of sequence complexity. However this is a statistical rather than a biophysical criterion, and some low complexity regions are structured, like collagen. This indicates that these models can confuse correlation and causation during training, and their representations should be interpreted carefully.
%        4. Furthermore, resolution of analyses limited because features calculated at level of regions, whose lengths can exceed 1,000 residues, rather than a fixed window around each residue.
%        5. Disorder scores are weakly correlated with many motifs. Furthermore, unclear if predictor is responding directly to these motifs as a signal of disorder, or they appear in regions which evolving towards disorder. Interestingly docking (DOC) and ligand binding (LIG) sites tend to have negative correlations whereas the modification (MOD) sites have positive correlations. Docking and ligand binding sites are generally mediated by small hydrophobic patches, so the predictor could be responding to increased hydrophobicity caused by additional of hydrophobic residue ``completing'' the motif rather than the motif itself. Likewise, phosphorylation sites are highly enriched in IDRs and many are targeted to disorder-promoting residues like serine or threonine. Thus, it's unclear if the predictor is responding directly to these features or features which are correlated with them. More in depths of the representations of these disorder predictors would shed led on these questions.
%    B. GO annotation enrichment analysis suggests disorder scores undergo rapid evolution in proteins involved in DNA repair and found in extracellular structure
%        1. Not corrected for multiple testing nor is the false discovery rate controlled, so should be considered as hypotheses for further investigation
%        2. However, this result is at least consistent with other known functions of disorder
%            i. Interactions with extracellular components and signaling, both highly enriched in disorder?
%            ii. These regions could contain MoRFs, so they are already on the boundary between folded and disorder
%            iii. As evolution optimizes these interactions, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles
%            iv. The variable scoring by the disorder predictor may reflect this kind of ``phrase transition''

>>>>

Despite the prevalence and importance of these IDRs in eukaryotic proteomes, associating them with specific molecular functions or biological processes remains a significant challenge. The sequences of IDRs are generally poorly conserved, so traditional bioinformatics approaches which rely on the conservation of amino acid sequences to identify homologous proteins and transfer annotations between them are largely unsuccessful when applied to IDRs. However, several recent studies have demonstrated evidence that IDRs are constrained to preserve ``distributed features'' such as flexibility~\cite{Daughdrill2007}, chemical composition~\cite{Moesa2012}, net charge~\cite{Zarin2017}, or charge distribution~\cite{Beh2012}. Because many sequences can yield a region with a specific composition, for example, this mode of constraint uncouples an IDR's fitness from its strict sequence of amino acid residues. Furthermore, in contrast to folded regions whose precise contacts and packing geometries are easily disrupted by amino acid substitutions, distributed features are robust to such changes as individual residues only weakly contribute to a region's fitness. For example, a mutation at one site in an IDR that changes its net charge is easily reversed by subsequent compensatory mutations elsewhere in the region. Thus, under this model the sequences of IDRs can rapidly diverge and still preserve their structural or functional properties.

%    B. Though other analyses have identified Though IDRs short linear motifs (SLiMs) are enriched in IDRs and are sometimes strongly conserved at specific positions, these residues constitute a small fraction of disordered regions. However, because these motifs are short and some are highly permissive at specific positions, they have high evolutionarily plasticity. Thus, motifs

%Though these motifs are sometimes conserved by position and can be identified as short segments of conservation in otherwise highly diverged alignments of IDRs, their brief lengths make them evolutionarily plastic. Thus, they are easily

@@

% III. Many of these analyses were limited to ... However, a recent study first to demonstrate patterns of selection in IDRs across an entire proteome using a set of 82 ``molecular features''

@@

%    E. Though subsequent studies have expanded on these findings by using techniques from machine learning and statistics the information from evolutionary signatures to predict the function of IDRs. and learn disorder-associated features in a self-supervised manner.

@@

% III. Disorder score analyses
%    A. The analysis of features and disorder scores is effectively a natural ``mutational scanning'' experiment [citation to Eddy?]
%        1. Contrasts of features and disorder scores shows the disorder predictor is sensitive to most features used in this analyses
%        2. However, highest correlations to expected features like fraction_disorder and hydropathy
%        3. Interestingly, negatively correlated with wf_complexity. Many disordered regions have low levels of sequence complexity. However this is a statistical rather than a biophysical criterion, and some low complexity regions are structured, like collagen. This indicates that these models can confuse correlation and causation during training, and their representations should be interpreted carefully.
%        4. Furthermore, resolution of analyses limited because features calculated at level of regions, whose lengths can exceed 1,000 residues, rather than a fixed window around each residue.
%        5. Disorder scores are weakly correlated with many motifs. Furthermore, unclear if predictor is responding directly to these motifs as a signal of disorder, or they appear in regions which evolving towards disorder. Interestingly docking (DOC) and ligand binding (LIG) sites tend to have negative correlations whereas the modification (MOD) sites have positive correlations. Docking and ligand binding sites are generally mediated by small hydrophobic patches, so the predictor could be responding to increased hydrophobicity caused by additional of hydrophobic residue ``completing'' the motif rather than the motif itself. Likewise, phosphorylation sites are highly enriched in IDRs and many are targeted to disorder-promoting residues like serine or threonine. Thus, it's unclear if the predictor is responding directly to these features or features which are correlated with them. More in depths of the representations of these disorder predictors would shed led on these questions.
%    B. GO annotation enrichment analysis suggests disorder scores undergo rapid evolution in proteins involved in DNA repair and found in extracellular structure
%        1. Not corrected for multiple testing nor is the false discovery rate controlled, so should be considered as hypotheses for further investigation
%        2. However, this result is at least consistent with other known functions of disorder
%            i. Interactions with extracellular components and signaling, both highly enriched in disorder?
%            ii. These regions could contain MoRFs, so they are already on the boundary between folded and disorder
%            iii. As evolution optimizes these interactions, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles
%            iv. The variable scoring by the disorder predictor may reflect this kind of ``phrase transition''
